### 2015-01-01 doc/thesis/system.tex
Moving the text about subtyping of unique and open table types back to the section about subtyping

There is a limitation on \emph{closed} table types that led us to
introduce \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Thus, we introduced \emph{unique} table types to represent the type of
the table constructor and avoid this limitation.

Moreover, we introduced \emph{open} table types to allow the refinement of table types,
an idiom very common in Lua.

Even though \emph{open} and \emph{unique} table types can interact with
\emph{closed} table types, they follow different subtyping rules.
We will present these rules in the next section along with the typing
rules that formalize the refinement of table types.
The presentation of these rules is easier in the next section,
as they include some unsafe behaviors that the refinement of
table types keeps track to make them safe.

@@

We also included two kinds of type cast in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
We will show that the first helps allowing safe aliasing between
table types, while the second allows the refinement of table types.
Due to the refinement of table types we also split variable names
into two categories: $n_{e}$ when they appear as expressions and
$n_{l}$ when they appear as left-hand values.

@@

The most unusual feature of our type system is the refinement of
table types.
The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
and \emph{unique} table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined a subtyping rule that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the type of
the value of the omitted key.
The following subtyping rule show this behavior:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
It is an auxiliary relation that defines the subtyping relation
on optional fields, and we defined it as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} v}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} \Const \; v}
\end{array}
\end{array}
\]

Although \emph{open} table types let programmers omit optional fields,
they do not let programmers initialize optional fields that are
mutable, unless they use the same type of the field in the initialization.
For instance,
\begin{align*}
\{ & ``firstname":\String,\\
   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{unique} table types to solve this problem.
We use \emph{unique} table types to denote the type of the
table constructor.
The following subtyping rule allow programmers to use table constructors
to initialize and omit optional fields:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
\end{array}
\]

The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
It is an auxiliary relation that makes \emph{unique} table types
have width and depth subtyping on all fields, and we defined it
as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\end{array}
\]

We restricted the refinement of table types to include only literal
keys because its purpose is to make it easier the construction of
table types that represent records.
For instance,
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$a[``y"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
uses field assignment to add fields $x$ and $y$ to the local $a$.
After these field assignments $a$ has type
$\{``x":\String, ``y":\Integer\}_{unique}$.

We do not allow the refinement of table types to change the type of
a field that is already present in the table type.
For instance,
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
do not type check, as we are trying to change the type of field
``x" from $\String$ to $\String \cup \Integer$.
More precisely, our type system does not allow the table type to
change towards a type that is not a subtype of the previous type.
In this example,
$\{``x":\String \cup \Integer\}_{unique} \not\subtype \{``x":\String\}_{unique}$.
We have this restriction because this kind of refinement could lead
to fields that have a too general type.

\textit{/* We need to close table types before assigning them to fields. */}
\begin{center}
\begin{tabular}{l}
$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in} \; a[``x"] \; {<}\{\}_{closed}{>} = \{\}$
\end{tabular}
\end{center}

>>>>

There is a limitation on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a closed table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of the tables that we known their contents.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
\end{array}
\]

Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
covariant keys.
We allow this behavior to \emph{unique} table types because we also
want to use them as a way to join table fields.
For instance, we may want to use the table constructor to initialize
a variable with a table type that describes a hash.

The rule \textsc{S-TABLE2} introduced the auxiliary relations
$\subtype_{u}$ and $\subtype_{o}$.
The first allows width and depth subtyping on all fields,
while the second allows the omission of optional fields.
We define them as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} v}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} \Const \; v}
\end{array}
\end{array}
\]

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\[
\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Even though \emph{unique} table types overcome our limitation,
its subtyping rule allows unsafe aliasing.
The covariance on mutable fields allows programs that create unsafe
aliases to pass the type checker.
For instance,
\begin{verbatim}
    local t1 = { foo = 5 }
    local t2:{"foo":integer?} = t1
    t2.foo = nil
\end{verbatim}
should not type check, as the last line is erasing the field \texttt{foo}
from \texttt{t1}.
This program would type checks as it is, because our implementation uses
local type inference to assign static types to unannotated locals.
In this example, the type of \texttt{t1} would be subtype of the type of \texttt{t2}.
More precisely,
\[
\{``foo":\Integer\}_{unique} \subtype \{``foo":\Integer \cup \Nil\}_{closed}
\]

In the next section we will show in more detail that this example
does not type check, because our type system promotes an alias
expression from \emph{unique} to \emph{closed}, and promotes the
aliased variable from \emph{unique} to \emph{open}.
This means that the previous example does not type check because
the type of the expression \texttt{t1} is not subtype of the type
of the variable \texttt{t2}.
More precisely,
\[
\{``foo":\Integer\}_{closed} \not\subtype \{``foo":\Integer \cup \Nil\}_{closed}
\]

Thus, we introduced \emph{open} table types to handle the aliasing of table types.
We define the subtyping relation among \emph{open} table types
and \emph{closed} or \emph{open} table types as follows:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

The rule \textsc{S-TABLE3} uses the same rules from \textsc{S-TABLE1}
for handling the type of the values of each key, and it uses the same
rules from \textsc{S-TABLE2} to allow joining fields plus omitting optional fields.

In the next section we will show in more detail how our type system
keeps track of \emph{open} and \emph{unique} table types to make them safe.
Moreover, we will also show that we can use them to allow the refinement
of table types.

@@

We also include two kinds of type cast in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
We will show that the first helps allowing safe aliasing between
table types, while the second allows the refinement of table types.
Due to the refinement of table types we also split variable names
into two categories: $n_{e}$ when they appear as expressions and
$n_{l}$ when they appear as left-hand values.

@@

\textit{/* \textbf{begin} TODO list */}

\begin{itemize}
\item Refinement with \_ENV
\item Assignment + left-hand side
\item Locals + inference rules + unique and closed table types
\end{itemize}

\textit{/* \textbf{end} TODO list */}

We start with the rule that defines the most unusual feature of
our type system: the refinment of table types.
The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
and \emph{unique} table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique}\\
       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i} \;\;\;
       \Nil \not\lesssim t_{1}
       \end{array}}
      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

\textit{/* Should we split into two rules instead of using $open|unique$? */}

We restricted the refinement of table types to include only literal
keys, because its purpose is to make it easier the construction of
table types that represent records.

We use the refinement of table types to handle the declaration of global variables.
In Lua, the assignment \texttt{v = 5} is syntactic sugar to \texttt{\string_ENV["v"] = 5}
when \texttt{v} is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an open table in the top-level scope.
For instance,
\begin{center}
\begin{tabular}{l}
$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
$\string_ENV[``y"] \; {<}\Integer{>} = 1$\\
\end{tabular}
\end{center}
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
We are assuming that $\string_ENV$ is in the environment and has type $\{\}_{open}$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

We do not allow the refinement of table types to change the type of
a field that is already present in the table type.
For instance,
\begin{center}
\begin{tabular}{l}
$\string_ENV[``x"] \; {<}\String{>} = ``foo";$\\
$\string_ENV[``x"] \; {<}\Integer{>} = 1$\\
\end{tabular}
\end{center}
do not type check, as we are trying to add a field that already exists
in $\string_ENV$.
We could allow the refinement of existing fields, but this could
lead to fields that have a too general type, which could decrease
the ammout of type checking.

\textit{/* We need to close table types before assigning them to fields. */}
\begin{center}
\begin{tabular}{l}
$\string_ENV[``x"] \; {<}\{\}_{unique}{>} = \{\}$
\end{tabular}
\end{center}
